Point-process based representation learning for Electronic Health Records

### 2023-02-16 Manuscript_v1.tex
update

\documentclass[journal,twoside,web]{ieeecolor}
\usepackage{generic}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}

@@

\usepackage{biblatex} %Imports biblatex package
% \addbibresource{sample.bib} %Import the bibliography file

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Point-process based representation learning for Electronic Health Records}
\begin{document}
\title{Point-process based representation learning for Electronic Health Records}
\author{Hojjat Karami, \IEEEmembership{Fellow, IEEE}, Anisoara Ionescu, and David Atienza, \IEEEmembership{Member, IEEE}
\thanks{This paragraph of the first footnote will contain the date on 
which you submitted your paper for review. It will also contain support 
information, including sponsor and financial support acknowledgment. For 
example, ``This work was supported by DigiPredict Grant BS123456.'' }
\thanks{H. Karami is with the EPFL, Lausanne, Switzerland (e-mail: hojjat.karami@epfl.ch). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
now with the Department of Physics, Colorado State University, Fort Collins, 
CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{T. C. Author is with 
the Electrical Engineering Department, University of Colorado, Boulder, CO 
80309 USA, on leave from the National Research Institute for Metals, 
Tsukuba, Japan (e-mail: author@nrim.go.jp).}}

@@

    Irregular @hornSetFunctionsTime2020 sampling of time series in electronic health records (EHRs) is challenging for model development. In addition, the pattern of missingness for certain clinical variables are not at random as it is determined by clinicians decision and the state of patient. Point process is a mathematical framework for handling event sequence data which is also consistent with the irregular sampling. We propose \emph{TEEDAM}, a deep neural network that can learn patient representation from irregulary sample time series as well as informative missingness pattern of certain laboratory variables. We performed various experiments to show the effectiveness of event and state encoding for characterization of conditional intensity functions as well as downstream prediction task. Results show that in some cases learning from patterns may improve the performance of prediction task.
\end{abstract}

\begin{IEEEkeywords}
    electronic health records (EHRs), Point Process, irregular sampling, informative missingness
\end{IEEEkeywords}

@@

healthcare systems are one of main area for deploying machine learning models. Thanks to recent advancements in data collection technologies, policies and computing power data are recorded in a hosptial. This data consists of multiple modalities frome tabular and time series to clinical notes and images. We focus on tabular and time series.

@@

conseq of irr sampling///

@@

imputation methods///

inherently compatible methods///

pp framework///

@@

\subsubsection{Temporal point process}
Consider an event sequence data  $\mathcal{D}=\{\mathcal{S}_{i}\}_{i=1}^N$ where each sample is represented as an event sequence  $\mathcal{S}_{i}=\{(t_i,e_i)\}_{j=1}^L$, where $L$ is number of event, $t_j$ is event's timestamp, and $e_j \in \mathbb{R}^M $ is the binary represntation of event marks (multi-class or multi-label). The history of events is denoted as $\mathcal{H}_t=\{(t_j,e_j):t_j<t \} $

In the multi-class setting where each mark can be exactly one of M marks, the point process is characterized by $M$ conidtional intensity functions (CIFs):

\begin{equation}\label{eq:11}
   \lambda_m^{*}(t)=\lim \frac{P(  \text{event of type m in } [t,t+\Delta t) |   \mathcal{H}_t  )}{\Delta t} 
\end{equation}

The default negative log-likelihood (NLL) of a point process for a single event sequnce $\mathcal{S}_{i}$ is defined as:

\begin{multline}\label{eq:11}
    -\log p(\mathcal{S}_{i})  =  - \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log \lambda_m^{*}(t_j) \\   
    + \sum_{m = 1}^{M} \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)
\end{multline}

If we assume that marks and events are independent,

\begin{multline}\label{eq:11}
    -\log p(\mathcal{S}_{i})  = \\
      - \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log p^{*}(e_j=m) \lambda^{*}(t_j) \\   
    + \sum_{m = 1}^{M} p^{*}(e_j=m) \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)
\end{multline}

[ntpp] has proposed the following CIF to extend for multi-label case:

\begin{multline}\label{eq:11}
    -\log p(\mathcal{S}_{i})  =  - \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log \lambda_m^{*}(t_j) \\   
    + \sum_{m = 1}^{M} \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)\\
    - \sum_{j = 1}^{L}\sum_{m = 1}^{M} (1-1(e_j=m)) \log \left(1-\lambda_m^{*}(t_j) \exp \left( - \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right) \right)
\end{multline}

@@

\subsubsection{Neural temporal point process}

Encoder-decoder architectures have proven to be effective in many applications. The main idea of a neural point process is to first encode the history of events until $t_j$ using different neural network architectures $h_j=Enc(\mathcal{H}_j)$. Then it tries to estimate $\lambda^*_m(t)$ or its integral for $ t_i < t \leq t_{i+1}$.

\subsubsection{Deep learning for irregular sampling}
An irregularly sampled data can be denoted as $\mathcal{D}=\{\mathcal{U}_i\}_{i=1}^{N}$ where $N$ is number of samples and each sample is represented as series of tuple $\mathcal{U}_i={(t_p,k_p,v_p)}$ where $t_p, k_p, v_p$ represents the time, modality and value of $p$-th datapoint respectively.

@@

 The key advantage of our proposed model is to combine a  transformer-based event encoder (TEE) with point process loss with a deep attention module (DAM) that can handle an irregularly sampled time series for a any downstream prediciton task. In this situation the data is represented as $\mathcal{D}=\{\mathcal{S}_i,\mathcal{U}_i\}_{i=1}^{N}$. General schematic of TEDAM is depicted in FIG.

@@

In the first step, we embed all event marks $E_{emb}=E \times W_emb$ where $E_i \in \mathbb{R}^{L \times M}$ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{M \times d_{emb} }  $ is the trainable embedding matrix.

In the second step, timestamps should be encoded and added to the event embedding, however, we propose to concatenate time encodings that can lead to better characterization of conditional intensity functions. Finally, the input of the transformer encoder will be $X_{emb}=[E_{emb}, T_{emb}] \in \mathbb{R}^{L \times (d_emb+t_emb)}$.

Here, we use the standard transformer encoder similar to [vaswani] with masking matrix to prevent the model from looking into the future. we obtain the encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $ where $h_j$ containes the all available information before occurence of $j$-th event.

@@

Similar to [setF], we use an attention-based aggregation approach for encoding all additional information. Each side information  $ (t_k, v_k, m_k) $ can be represented by $s_k=(z(t_k), v_k, m_k)$. we define attention $a(\mathcal{S}_k,s_k )$

We define $\mathcal{S}_p$ to be the set of the first $p$ available information.
The goal is to calculate $a(\mathcal{S}_p,s_k ), k \leq p$ that is the relevance of $k$-th observation $s_k$ to the first $p$ observed values $\mathcal{S}_p$.
This is achieved by computing an embedding of the set elements using a smaller set functions $f^{\prime}$, and projecting the concatenation of the set representation and the individual set element into d-dimensional space:

The goal is to calculate $a(\mathcal{S}_p,s_k ), k \leq p$ that is the relevance of $k$-th observation $s_k$ to the first $p$ observed values $\mathcal{S}_p$.

This is achieved by embedding set elements $\mathcal{S}_p$ using a simpler set function $f^{\prime}$ such as a cumulative mean.

@@

\begin{equation*}
    Z^{red}_i=Z_p where p=argmax(z \leq i)
\end{equation*}

@@

All formulas are:

\begin{equation*}
    \begin{cases}

f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(   
 \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) 
\right) \\

K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K \\
e_p=\frac{K_p.w^q}{\sqrt{d}  }\\

a(\mathcal{S}_p,s_k)=\frac{\exp (e_p)}{\sum_{k\leq p} \exp (e_k)  } \\

f(\mathcal{S}_p) = \sum_{s_k \in \mathcal{S}_p}  a(\mathcal{S}_p,s_k)h_{\theta}(s_k) \\

z_p=g_{\psi}\left(         f(\mathcal{S}_p)         \right) \\

\end{cases} 
\end{equation*}

\begin{equation}\label{eq:1}
    [z(t_j)]_{k} = 
     \begin{cases} 
        \sin\left(  \frac{t_j}{\mathcal{T}^{(k-1)/d_t} }  \right) & \text{if } k \text{ is odd}\\
        \sin\left(  \frac{t_j}{\mathcal{T}^{k/d_t} }  \right) & \text{if } k \text{ is even}

        \end{cases} 
\end{equation} 
\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+\mathbf{y}_k^T \mathbf{s}_{enc}(t_j) +b_k   \right)
\end{equation}

\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+b_k   \right)
\end{equation}

@@

\begin{multline*}
    \lambda_m(t)=softplus(\mu_{m,i+1}+\\
    (\eta_{m,i+1}- \mu_{m,i+1}) \exp(-\gamma_{m,i+1}(t-t_i))    ),
\end{multline*}

@@

\textbf{Physionet 2012 Mortality Prediction Challenge (\emph{P19})}.
The 2012 Physionet challenge dataset (Goldberger et al., 2000),contains 12, 000 ICU stays each of which lasts at least 48 h. For each stay, a set of general descriptors (such as gender or age) are collected at admission time. Depending on the course of the stay and patient status, up to 37 time series variables were measured (e.g. blood pressure, lactate, and respiration rate). While some modalities might be measured in regular time intervals (e.g. hourly or daily), some are only collected when required; moreover, not all variables are available for each stay.

@@

effect of time concatenation

compare single+mark with mc or ml

% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table*}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{cccccc}
      \toprule
            &       & \multicolumn{2}{c}{TEEDAM} &       &  \\
  \cmidrule{3-4}    Dataset & Metric & concat & sum   & SAHP  & GRU-CP \\
      \midrule
      \multirow{2}[2]{*}{SO} & LL/\#events &       &       &       &  \\
            & AUROC &       &       &       &  \\
      \midrule
      \multirow{2}[2]{*}{Synthea} & LL/\#events &       &       &       &  \\
            & AUROC &       &       &       & 0.85(.014) \\
      \midrule
      \multirow{2}[2]{*}{ReTweet} & LL/\#events & 1.83 (0.09) & 1.5 (0.18) &       &  \\
            & AUROC & 0.68 (0.03) & 0.67 (0.05) &       & 0.611(0.001) \\
      \bottomrule
      \end{tabular}%
    \label{tab:addlabel}%
  \end{table*}%

@@

Table 1 shows the result for estimation of negative likelihood in different datasets and scnearios. It is obvious that state encoding has led to lower NLL.

% Table generated by Excel2LaTeX from sheet 'state enc'
\begin{table}[h!]
    \centering
    \caption{Add caption}
      \begin{tabular}{ccccc}
      \toprule
            &       & \multicolumn{3}{c}{Model} \\
  \cmidrule{3-5}    Dataset & setting & TE    & TE+DAM & TE+noise \\
      \midrule
      \multirow{3}[2]{*}{P12} & sc    & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & mc1   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & mc2   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
      \midrule
      \multirow{3}[2]{*}{p19} & sc    & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & mc1   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & mc2   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
      \bottomrule
      \end{tabular}%
    \label{tab:addlabel}%
  \end{table}%

@@

\subsection{Downstream task with event encoding}

Another key element of our work is to show the effectiveness of point process modeling for a down-stream task. In Table, we have reported the performance metrics for the mortality prediction task. We have compared our results with several sota's DL models that are compatible with irregular time series.

@@

\begin{thebibliography}{00}

\bibitem{b1} G. O. Young, ``Synthetic structure of industrial plastics,'' in \emph{Plastics,} 2\textsuperscript{nd} ed., vol. 3, J. Peters, Ed. New York, NY, USA: McGraw-Hill, 1964, pp. 15--64.

\bibitem{b2} W.-K. Chen, \emph{Linear Networks and Systems.} Belmont, CA, USA: Wadsworth, 1993, pp. 123--135.

\bibitem{b3} J. U. Duncombe, ``Infrared navigation---Part I: An assessment of feasibility,'' \emph{IEEE Trans. Electron Devices}, vol. ED-11, no. 1, pp. 34--39, Jan. 1959, 10.1109/TED.2016.2628402.

\bibitem{b4} E. P. Wigner, ``Theory of traveling-wave optical laser,'' \emph{Phys. Rev}., vol. 134, pp. A635--A646, Dec. 1965.

\bibitem{b5} E. H. Miller, ``A note on reflector arrays,'' \emph{IEEE Trans. Antennas Propagat}., to be published.

>>>>

\documentclass[journal,twoside,web]{ieeecolor}
\usepackage{generic}
% \usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}

@@

\usepackage{biblatex} %Imports biblatex package
\addbibresource{zotero.bib} %Import the bibliography file

% \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Point-process based representation learning for Electronic Health Records}
\begin{document}
\title{Point-process based representation learning for Electronic Health Records}
\author{Hojjat Karami, \IEEEmembership{Fellow, IEEE}, Anisoara Ionescu, and David Atienza, \IEEEmembership{Member, IEEE}
\thanks{This paragraph of the first footnote will contain the date on 
which you submitted your paper for review. It will also contain support 
information, including sponsor and financial support acknowledgment. For 
example, ``This work was supported by DigiPredict Grant BS123456.'' }
\thanks{H. Karami is with the EPFL, Lausanne, Switzerland (e-mail: hojjat.karami@epfl.ch). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
now with the Department of Physics, Colorado State University, Fort Collins, 
CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{T. C. Author is with 
the Electrical Engineering Department, University of Colorado, Boulder, CO 
80309 USA, on leave from the National Research Institute for Metals, 
Tsukuba, Japan (e-mail: author@nrim.go.jp).}}

@@

    Irregular  sampling of time series in electronic health records (EHRs) is challenging for model development. In addition, the pattern of missingness for certain clinical variables are not at random as it is determined by clinicians decision and the state of patient. Point process is a mathematical framework for handling event sequence data which is also consistent with the irregular sampling. We propose \emph{TEEDAM}, a deep neural network that can learn patient representation from irregulary sample time series as well as informative missingness pattern of certain laboratory variables. We performed various experiments to show the effectiveness of event and state encoding for characterization of conditional intensity functions as well as downstream prediction task. Results show that in some cases learning from patterns may improve the performance of prediction task.
\end{abstract}

\begin{IEEEkeywords}
    electronic health records (EHRs), Point Process, irregular sampling, informative missingness,
\end{IEEEkeywords}

@@

healthcare systems are one of main area for deploying machine learning models. Thanks to recent advancements in data collection technologies, policies and computing power data are recorded in a hosptial. This data consists of multiple modalities frome tabular and time series to clinical notes and images. We focus on tabular and time series. \cite{zuoTransformerHawkesProcess2020a}

@@

conseq of irr sampling/// Irregularly sampled time series in electronic health records can have serious consequences for data analysis and decision making. Irregular sampling occurs when the time intervals between data points are inconsistent, which can lead to biased or inaccurate results when trying to make inferences about the underlying process. This can also make it difficult to accurately compare the data over time, as well as to detect patterns or trends. Furthermore, irregular sampling can cause problems for machine learning algorithms, which may struggle to accurately capture the relationships between variables in the data. This can result in poor predictions or incorrect diagnoses, potentially leading to negative impacts on patient care. To mitigate these consequences, it is important to ensure that electronic health records are sampled at regular intervals, or that appropriate statistical techniques are applied to account for the irregularity.

@@

imputation methods/// Imputation techniques are widely used in electronic health records (EHRs) to handle missing data. EHRs often contain incomplete or missing information due to various reasons such as measurement errors, system failures, or data entry issues. Imputation refers to the process of filling in the missing data with plausible values based on the available information. There are several imputation techniques available, each with their own strengths and weaknesses. For example, mean imputation replaces missing values with the mean of the available values in the same column, while regression imputation uses a regression model to estimate the missing values based on the relationship between the variables. Other popular imputation methods include multiple imputation, hot deck imputation, and k-nearest neighbors imputation. The choice of imputation technique depends on the type and amount of missing data, as well as the goals of the analysis. Regardless of the technique used, it is important to carefully consider the impact of imputed values on the analysis and to report the imputation method used in the results.

inherently compatible methods/// Deep learning models that are compatible with irregularly sampled time series are important in many real-world applications, such as stock market predictions, speech recognition, and medical diagnosis. These models need to be able to handle data that is not evenly spaced, as is often the case in time-sensitive applications. One popular deep learning architecture that can handle irregularly sampled time series is the Recurrent Neural Network (RNN). RNNs are well-suited for this task because they have the ability to process sequential data, taking into account not only the current input but also the previous inputs. Another architecture that can handle irregularly sampled time series is the Convolutional Neural Network (CNN), which can be used to extract features from the data and then pass the processed data to an RNN for further analysis. Additionally, Attention Mechanisms can be integrated into RNNs and CNNs to help the model focus on important features in the data. Overall, these deep learning models offer a powerful toolset for handling irregularly sampled time series and provide useful insights into this type of data.

pp framework/// Point processes are mathematical models used to describe the distribution of events in time or space. These events can be occurrences of earthquakes, nerve impulses, customer purchases, or anything else that can be counted or measured. Neural Point Processes are a type of machine learning models that learn the patterns and relationships between events using artificial neural networks. These models can be used for tasks such as predicting future events, estimating the rate of event occurrence, or identifying correlations between events. They offer a flexible and powerful way to analyze point process data, as they can handle complex dependencies between events and incorporate prior knowledge about the process.

@@

\subsubsection*{Temporal point process}

Consider an event sequence data  $\mathcal{D}=\{\mathcal{S}_{i}\}_{i=1}^N$ where each sample is represented as an event sequence  $\mathcal{S}_{i}=\{(t_i,e_i)\}_{j=1}^L$, where $L$ is the total number of events, $t_j$ is event's timestamp, and $e_j \in \mathbb{R}^M $ is the binary represntation of event marks (multi-class or multi-label). The history of events is denoted as $\mathcal{H}_t=\{(t_j,e_j):t_j<t \}$.

The core idea of the point process framework is the definition of conditional intensity functions (CIFs) which is the probability of the occurence of an event of type $m$ in an infinitesimal time window $[t,t+dt)$:

\begin{equation} 
   \lambda_m^{*}(t)=\lim \frac{P(  \text{event of type m in } [t,t+\Delta t) |   \mathcal{H}_t  )}{\Delta t} 
\end{equation}

Multivariate Hawkes process is the traditional approach to characterize CIFs:

\begin{equation} 
    \lambda_m^{*}(t)= \mu_m + \sum_{(t^{\prime},e^{\prime})\in \mathcal{H}_t} \phi(t-t^{\prime})
 \end{equation}

\subsubsection*{Neural temporal point process}

Encoder-decoder architectures have proven to be effective in many applications. The main idea of a neural temporal point process (NTPP) is to first encode the history of events until $t_j$ using a neural network architecture $h_j=Enc(\mathcal{H}_{j+1};\theta)$. Then it tries to estimate $\lambda^*_m(t)=Dec(h_j;\phi)$ for $ t \in (t_j,t_{j+1}]$.

Initial works used recurrent neural networks either in a discrete way [16], a continuous way [39], or a multi-scale multi-channel architecture [24]; then self-attention was also used by [54, 57]. [57] develop a Structured Transformer Hawkes (STH) model that studies the relationship between point processes. It assumes the modeling of one point process might contribute to the modeling of another. Each vertex in the STH’s graph is associated with a point process. Other efforts have been made to model the cumulative intensity function [42] and conditional probability density [46]. One issue that has been ignored in these works is that these NPP models take all types of historic events to predict the future ones, neglecting that some event types may not contribute to the prediction of another type. This study aims to correct this defect by learning to omit those non-contributing event types when predicting the target type via NPPs.

\subsubsection*{Parameter Estimation}

Based on conditional intensity function, it is straightforward to derive conditional probability density function $p^{*}_{m}(t)$ in $(t_j, t_{j+1}]$:

\begin{equation} 
    p_m^{*}(t)=\lambda_m^{*}(t) \exp \left[-\sum_{m=1}^{M} \int_{t_j}^{t_{j+1}}\lambda_m^{*}(t^{\prime})dt^{\prime}\right]
 \end{equation}

 Intuitively, the integral term indicates that there is no event of any type in the interval.

In the multi-class setting, the log-likelihood (LL) of a point process for a single event sequnce $\mathcal{S}_{i}$ is defined as:

\begin{multline} 
    \log p_{mc}(\mathcal{S}_{i})  =  \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log p_m^{*}(t_j) \\   
    = \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log \lambda_m^{*}(t_j)\\
    \sum_{m = 1}^{M} \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)
\end{multline}

However, in many cases such as EHRs it is common to have co-occuring events. [ntpp] proposed to add:

% If we assume that marks and events are independent,

% \begin{multline} 
%     -\log p(\mathcal{S}_{i})  = \\
%       - \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log p^{*}(e_j=m) \lambda^{*}(t_j) \\   
%     + \sum_{m = 1}^{M} p^{*}(e_j=m) \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)
% \end{multline}

\begin{multline} 
    \log p_{ml}(\mathcal{S}_{i}) = \log p_{mc}(\mathcal{S}_{i}) \\
    + \sum_{m = 1}^{M} (1-1(e_j=m)) \log \left(1-p_m^{*}(t_j)  \right)
\end{multline}

% \begin{multline} 
%     -\log p(\mathcal{S}_{i})  =  - \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log \lambda_m^{*}(t_j) \\   
%     + \sum_{m = 1}^{M} \left(   \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right)\\
%     - \sum_{j = 1}^{L}\sum_{m = 1}^{M} (1-1(e_j=m)) \log \left(1-\lambda_m^{*}(t_j) \exp \left( - \int_{0}^{T}  \lambda_m^{*}(s) \,ds  \right) \right)
% \end{multline}

@@

Another approach is the marked case which assumes that marks and timestamps are conditionally independent given $\mathcal{H}_t$.

\begin{multline} 
    \log p_{marked}(\mathcal{S}_{i}) =    
    \sum_{j = 1}^{L}\sum_{m = 1}^{M} 1(e_j=m)   \log p^{*}(e_j=m) \\
    +\sum_{j = 1}^{L}\lambda^{*}(t_j) - \int_{t_0}^{t_{L}}\lambda^{*}(t^{\prime})dt^{\prime}  
\end{multline}

\subsubsection*{Deep learning for irregular sampling}
An irregularly sampled data can be denoted as $\mathcal{D}=\{\mathcal{U}_i\}_{i=1}^{N}$ where $N$ is number of samples and each sample is represented as series of tuple $\mathcal{U}_i={(t_p,k_p,v_p)}$ where $t_p, k_p, v_p$ represents the time, modality and value of $p$-th datapoint respectively.

@@

 The key advantage of our proposed model is to combine a  transformer-based event encoder (TEE) with a deep attention module (DAM) that can handle an irregularly sampled time series for a any downstream prediciton task. In this case the data is represented as $\mathcal{D}=\{\mathcal{S}_i,\mathcal{U}_i\}_{i=1}^{N}$. General schematic of TEDAM is depicted in FIG.

@@

In the first step, we embed all event marks $E_{emb}=E \times W_{emb}$ where $E_i \in \mathbb{R}^{L \times M}$ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{M \times d_{emb} }  $ is the trainable embedding matrix.

In the second step, timestamps should be encoded and added to the event embedding, however, we propose to concatenate time encodings that can lead to better characterization of conditional intensity functions. Finally, the input of the transformer encoder will be $X_{emb}=[E_{emb}, T_{emb}] \in \mathbb{R}^{L \times (d_{emb}+t_{emb})}$.

Here, we use the standard transformer encoder similar to [vaswani] with masking matrix to prevent the model from looking into the future. we obtain the encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $ where $h_j$ containes the all available information until occurence of $j$-th event.

@@

Similar to [setF], we use an attention-based aggregation approach for encoding all additional information. Each side information  $ (t_k, v_k, m_k) $ can be represented by $s_k=(z(t_k), v_k, m_k)$. we define attention $a(\mathcal{U}_k,s_k )$

We define $\mathcal{U}_p$ to be the set of the first $p$ available information.
The goal is to calculate $a(\mathcal{U}_p,s_k ), k \leq p$ that is the relevance of $k$-th observation $s_k$ to the first $p$ observed values $\mathcal{U}_p$.
This is achieved by computing an embedding of the set elements using a smaller set functions $f^{\prime}$, and projecting the concatenation of the set representation and the individual set element into d-dimensional space:

@@

\begin{equation*}
    Z^{red}_i=Z_p \text{ where } p=argmax(z \leq i)
\end{equation*}

@@

\subsection{All formulas}

% \begin{equation*}
%     \begin{cases}

% f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(   
%  \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) 
% \right) \\

% K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K \\
% e_p=\frac{K_p.w^q}{\sqrt{d}  }\\

% a(\mathcal{S}_p,s_k)=\frac{\exp (e_p)}{\sum_{k\leq p} \exp (e_k)  } \\

% f(\mathcal{S}_p) = \sum_{s_k \in \mathcal{S}_p}  a(\mathcal{S}_p,s_k)h_{\theta}(s_k) \\

% z_p=g_{\psi}\left(         f(\mathcal{S}_p)         \right) \\

% \end{cases} 
% \end{equation*}

% \begin{equation}\label{eq:1}
%     [z(t_j)]_{k} = 
%      \begin{cases} 
%         \sin\left(  \frac{t_j}{\mathcal{T}^{(k-1)/d_t} }  \right) & \text{if } k \text{ is odd}\\
%         \sin\left(  \frac{t_j}{\mathcal{T}^{k/d_t} }  \right) & \text{if } k \text{ is even}

%         \end{cases} 
% \end{equation} 
% \begin{equation}
%     \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+\mathbf{y}_k^T \mathbf{s}_{enc}(t_j) +b_k   \right)
% \end{equation}

% \begin{equation}
%     \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+b_k   \right)
% \end{equation}

@@

\begin{multline*}
    \lambda_m(t)=\text{softplus}(\mu_{m,i+1}+\\
    (\eta_{m,i+1}- \mu_{m,i+1}) \exp(-\gamma_{m,i+1}(t-t_i))    ),
\end{multline*}

@@

\textbf{Physionet 2012 Mortality Prediction Challenge (\emph{P12})}.
The 2012 Physionet challenge dataset (Goldberger et al., 2000),contains 12, 000 ICU stays each of which lasts at least 48 h. For each stay, a set of general descriptors (such as gender or age) are collected at admission time. Depending on the course of the stay and patient status, up to 37 time series variables were measured (e.g. blood pressure, lactate, and respiration rate). While some modalities might be measured in regular time intervals (e.g. hourly or daily), some are only collected when required; moreover, not all variables are available for each stay.

@@

[TB1] shows the performance metrics of different datasets.

In general, time concatenation has lead to better F1-score/AUROC as well as LL/events compared to adding time. This idea can be used for future research in NTPPs for better characterization of CIFs.

Another interesting fact is that in some datasets (A, B) next mark prediction has competetive performance compared to the point process loss. However, in many cases, point process loss significantly improves the performance metrics which indicates that modeling non-event likelihood could be beneficial. In addition, we propose to use next-mark prediction case as a baseline.

Another interesting results is that independent assumption in the marked loss may lead to worse performance compared to MC/ML loss.

% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table*}[htbp]
  \centering
  \caption{Add caption}
    \begin{tabular}{cccccccccccccc}
    \toprule
          &       & \multicolumn{8}{c}{TEEDAM}                                    &       &       &       &  \\
\cmidrule{3-10}          &       & \multicolumn{2}{c}{AE (next mark)} &       & \multicolumn{2}{c}{PP(single+mark)} &       & \multicolumn{2}{c}{PP (MC/ML)} &       &       &       &  \\
\cmidrule{3-4}\cmidrule{6-7}\cmidrule{9-10}    Dataset & Metric & concat & sum   &       & concat & sum   &       & concat & sum   & Latent & SAHP  & THP   & GRU-CP \\
    \midrule
    \multirow{2}[2]{*}{SO} & LL/\#events & ND    & ND    &       & -0.56 & -0.57 &       &       &       & -1.54 & -1.86 & -1.84 & NR \\
          & F1-score & 27.86 & 28.79 &       & 29.10 & 28.95 &       &       &       & 28.34(0.19) & 24.12 & 23.89 & 26 \\
    \midrule
    \multirow{2}[2]{*}{ReTweet (MC)} & LL/\#events & ND    & ND    &       &       &       &       &       &       & -3.89 & -4.56 & -4.57 & NR \\
          & F1-score &       &       &       &       &       &       &       &       & 58.29 & 53.92 & 53.86 & NR \\
    \midrule
    \multirow{2}[2]{*}{Synthea} & LL/\#events & ND    & ND    &       &       &       &       &       &       & ND    & ND    & ND    & NR \\
          & AUROC &       &       &       &       &       &       &       &       & ND    & ND    & ND    & 0.85(.014) \\
    \midrule
    \multirow{2}[2]{*}{ReTweet (ML)} & LL/\#events & ND    & ND    &       & 1.86  & 1.4   &       &       &       & ND    & ND    & ND    & NR \\
          & AUROC & 63.3  & 61.01 &       & 68.7  & 66.2  &       & 73.87 &       & ND    & ND    & ND    & 0.611(0.001) \\
    \bottomrule
    \end{tabular}%
  \label{tab:addlabel}%
\end{table*}%

@@

[TB2] shows the result for estimation of negative likelihood for the two datasets.

In general we can see that using state information results in lower NLL, however, it cannot neccessarily increase the AUROC for next time prediction.

Another observation is that more data will lead to higher LL.

% Table generated by Excel2LaTeX from sheet 'state enc'
\begin{table}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{ccccc}
      \toprule
            &       & \multicolumn{3}{c}{Model} \\
  \cmidrule{3-5}    Dataset & setting & TE    & TE+DAM & TE+noise \\
      \midrule
      \multirow{3}[2]{*}{P12} & sc    & -1.99 & -1.77 & -1.92 \\
            & mc1   & -0.99 & -1.14 & -0.89 \\
            & mc2   & -1.5  & -1.33 & -1.47 \\
      \midrule
      \multirow{3}[2]{*}{P19} & sc    & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & mc1   & -1.81 & -1.62 & -1.69 \\
            & mc2   & -2.77 & -2.38 & -2.67 \\
      \bottomrule
      \end{tabular}%
    \label{tab:addlabel}%
  \end{table}%

@@

\subsection{Downstream task}

[TB3] indicates the result for Mortality/sepsis prediction in different settings and hospital centers.

In some cases, incorporating event embeddings can lead to better performance.

@@

% \begin{thebibliography}{00}

% \bibitem{b1} G. O. Young, ``Synthetic structure of industrial plastics,'' in \emph{Plastics,} 2\textsuperscript{nd} ed., vol. 3, J. Peters, Ed. New York, NY, USA: McGraw-Hill, 1964, pp. 15--64.

% \bibitem{b2} W.-K. Chen, \emph{Linear Networks and Systems.} Belmont, CA, USA: Wadsworth, 1993, pp. 123--135.

% \bibitem{b3} J. U. Duncombe, ``Infrared navigation---Part I: An assessment of feasibility,'' \emph{IEEE Trans. Electron Devices}, vol. ED-11, no. 1, pp. 34--39, Jan. 1959, 10.1109/TED.2016.2628402.

% \bibitem{b4} E. P. Wigner, ``Theory of traveling-wave optical laser,'' \emph{Phys. Rev}., vol. 134, pp. A635--A646, Dec. 1965.

% \bibitem{b5} E. H. Miller, ``A note on reflector arrays,'' \emph{IEEE Trans. Antennas Propagat}., to be published.

% \bibitem{b6} E. E. Reber, R. L. Michell, and C. J. Carter, ``Oxygen absorption in the earth's atmosphere,'' Aerospace Corp., Los Angeles, CA, USA, Tech. Rep. TR-0200 (4230-46)-3, Nov. 1988.
